---
title: "VolcanoCO2"
author: "Daryl Hegyi"
date: "Friday, March 06, 2015"
output: 
  html_document:
    keep_md: true
---
## The Contribution of Volcanoes to Global Warming.
Given a continuous long-duration record of CO2 at a single location (Mauna Loa), and a list of volcano eruptions during the monitored period, determine the amount of CO2 injected into the atmosphere by each volcano, and use that information to quantify the CO2 that can be attributed to geological (volcanic) processes.


```{r Setup, echo=FALSE}
library(scales)
library(lubridate)
library(pspline)
```
CO2 data: <file://./data/co2_mm_mlo.txt>
```{r ReadData}
co2 <- read.table("./data/co2_mm_mlo.txt", header=TRUE, quote="\"", na.strings = "-99.99")
names(co2)<-c("year", "month", "decimal_date", "average", "interpolated", "trend", "numdays")
summary(co2)
```

* Loading up the Volcanos dataset
```{r LoadVolcanos, echo=FALSE}

vol.Classes <- c( "factor", "Date", "Date", "numeric")

vol <- read.csv("./data/Volcanos.csv", na.strings = "NA", colClasses=vol.Classes)

vol$erupted<-decimal_date(as.Date(vol$Eruption_date))

#fixup the NA/s in average by using the interpolated values
co2.na<-is.na(co2$average)
co2$average[co2.na] <- co2$interpolated[co2.na]

#vol$enderupt[is.na(vol$enderupt)]<-vol$Eruption_date
#vol$enderupt<-decimal_date(as.Date(vol$EndEruption_date))
```
* Set up plot functions.
```{r PlotFunctions}
plot.volcanos<- function(vol.col="green",...) abline(v=vol$erupted, col=vol.col)
plot.co2<- function(x=co2$decimal_date,...) {plot(x, xlab="Date", ...); plot.volcanos()}
lines.co2<- function(...) {lines(x=co2$decimal_date, ...)}
```

* Looking at the data - blue is the monthly average, and red is the trend
```{r FirstPlot, echo=FALSE}
plot.co2(y=co2$average, type="l",  main="Average Monthly CO2 at Mauna Loa", 
					  ylab="CO2 (ppm)", col="blue")
lines.co2(y=co2$trend, type="l", col="black")


```

* I'll just use the trend data, as they have nicely cleaned it for me, removing the seasonal variation.
* I'll try to fit a couple of polynomials - linear, square, and cube.
```{r CurveFit}

x<-co2$decimal_date
y<-co2$trend
#fit first degree polynomial equation:
fit1  <- lm(y~x)
#second degree
fit2 <- lm(y~poly(x,2,raw=TRUE))
fit3 <- lm(y~poly(x,3,raw=TRUE))

op<-par(mfrow = c(2, 2))
plot(fit1, main="fit 1 (linear)")
plot(fit2, main="fit 2 (square)")
plot(fit3, main="fit 3 (cubic)")
par(op)

```

* I'll look at the residuals. 
```{r Residuals}
# now, look at the residuals

plot.co2(resid(fit1), col=alpha("red",.8), type="l", lwd=8, main="Residuals")
lines.co2(resid(fit2), col=alpha("blue",.5), type="l", lwd=5)
lines.co2(resid(fit3), col=alpha("yellow",.3), type="l", lwd=3)
```

* There doesn't seem to be much difference between the square and cube fits
 we will just use the square fit.
* I would like to subract the residuals, and perform a new fit, to see if the data cleans up anymore
* remove the residuals from the original data, then re-run the simulation
```{r, SubResiduals}
op<-par(mfrow = c(2, 2))

trr<- co2$trend - fit2$residuals
fit2b <- lm(trr~poly(x,2,raw=TRUE))
plot(fit2b, main="Fit2b (square)")

trr<- co2$trend - fit3$residuals
fit3b <- lm(trr~poly(x,3,raw=TRUE))
plot(fit3b, main="Fit3b (cubic)")

fit4 <- lm(trr~poly(x,4,raw=TRUE))
plot.co2(log(fit4$residuals+1e-12), main="Fit4 (quad)", type="l")

par(op)

plot.co2(y=log(resid(fit2)+1.56), type="l")
#lines(x=co2$decimal_date, y=log(rp2+1.6), col=alpha("red",.5))

plot.co2( y=co2$trend, type="l", lwd=8, col="blue")
lines.co2(y=co2$trend-resid(fit2), type="l", lw=5, col="green")
lines.co2(y=co2$trend-resid(fit2b), type="l", lw=3, col="red")

```

* See if there is any significance to square or cubic fit
```{r FitTest}
anova(fit1,fit2)
anova(fit1,fit3)

```

* That looks like some real activity. There is definitely a response to the eruptions, but it's subtle.
## try a spline fit to the residuals
```{r DiffEqs}
dds <- sm.spline(x, resid(fit2))
summary(dds)
ddyb <- predict(dds, x, 1)
summary(ddyb)
#plot(x, yp2, type='l', col="red")
plot.co2(y=ddyb, type='l', col="blue")

plot.co2(y=log(ddyb+4), type='l', col="blue")
```
## Use VAR 
```{r vars, echo=FALSE	}
library(vars)

var.co2 <- VAR(co2[,c("decimal_date", "trend")], p = 2, type = "const", )

op<-par(mfrow=c(2,2))
plot(var.co2)
par(mrows=c(2,1))
plot.co2( main="Trend Residuals",
	x=var.co2$varresult$decimal_date$fitted.values, 
	y=var.co2$varresult$trend$residuals, type="l")

plot.co2( main="log(Trend Residuals)",
	x=var.co2$varresult$decimal_date$fitted.values, 
	y=log(var.co2$varresult$trend$residuals+1), type="l")

#plot(irf(var.co2, impulse = "e", response = c("prod", "rw", "U"), boot = T))
par(op)
```

## Zoom in on the volcano timeframes
Showing the CO2 Residuals from 1/2 year before until 2 years after the event
```{r ZoomVolcanos}
par(mfrow=c(3,2))
for (date.erupted in vol$erupted) 
{
	plot.co2( main= "Trend Residuals", ylab="CO2 Res",
		x=var.co2$varresult$decimal_date$fitted.values, 
		y=var.co2$varresult$trend$residuals, type="l",
		xlim=c(date.erupted-0.5, date.erupted+2) )

	plot.co2( main="log(Trend Residuals)", ylab="CO2 Res",
		x=var.co2$varresult$decimal_date$fitted.values, 
		y=log(var.co2$varresult$trend$residuals+1), type="l",
		xlim=c(date.erupted-.5, date.erupted+2) )
}
```