---
title: "VolcanoCO2"
author: "Daryl Hegyi"
date: "Friday, March 06, 2015"
output: 
  html_document:
    keep_md: true
---
## The Contribution of Volcanoes to Global Warming.
Given a continuous long-duration record of CO2 at a single location (Mauna Loa), and a list of volcano eruptions during the monitored period, determine the amount of CO2 injected into the atmosphere by each volcano, and use that information to quantify the CO2 that can be attributed to geological (volcanic) processes.


```{r Setup, echo=FALSE}
library(scales)
library(lubridate)
library(pspline)
```
CO2 data: <http://co2now.org/Current-CO2/CO2-Now/scripps-co2-data-mauna-loa-observatory.html>
```{r ReadData}
co2 <- read.table("./data/co2_mm_mlo.txt", header=TRUE, quote="\"", na.strings = "-99.99")
names(co2)<-c("year", "month", "decimal_date", "average", "interpolated", "trend", "numdays")
summary(co2)
```

* Loading up the Volcanos dataset
```{r LoadVolcanos, echo=FALSE}

vol.Classes <- c( "factor", "Date", "Date", "numeric")

vol <- read.csv("./data/Volcanos.csv", na.strings = "NA", colClasses=vol.Classes)

vol$erupted<-decimal_date(as.Date(vol$Eruption_date))
#vol$enderupt[is.na(vol$enderupt)]<-vol$Eruption_date
#vol$enderupt<-decimal_date(as.Date(vol$EndEruption_date))
```
* Set up plot functions.
```{r PlotFunctions}
plot.volcanos<- function(...) abline(v=vol$erupted)
plot.co2<- function(...) {plot(x=co2$decimal_date, xlab="Date", ...); plot.volcanos()}
lines.co2<- function(...) {lines(x=co2$decimal_date, ...)}
```

* Looking at the data - blue is the monthly average, and red is the trend
```{r FirstPlot, echo=FALSE}
plot.co2(y=co2$average, type="l",  main="Average Monthly CO2 at Mauna Loa", 
					  ylab="CO2 (ppm)", col="blue")
lines.co2(y=co2$trend, type="l", col="black")


```

* I'll just use the trend data, as they have nicely cleaned it for me, removing the seasonal variation.
* I'll try to fit a couple of polynomials - linear, square, and cube.
```{r CurveFit}

x<-co2$decimal_date
y<-co2$trend
#fit first degree polynomial equation:
fit1  <- lm(y~x)
#second degree
fit2 <- lm(y~poly(x,2,raw=TRUE))
fit3 <- lm(y~poly(x,3,raw=TRUE))

par(mfrow = c(2, 2))
plot(fit1, main="fit 1 (linear)")
plot(fit2, main="fit 2 (square)")
plot(fit3, main="fit 3 (cubic)")
par(mfrow = c(1,1))
xx <- seq(1958.00,2016.00, by=0.66)
plot(x,y,pch=19,ylim=c(310,410),type="l")
lines(xx, predict(fit1, data.frame(x=xx)), col="red")
lines(xx, predict(fit2, data.frame(x=xx)), col="green")
lines(xx, predict(fit3, data.frame(x=xx)), col="blue")


```

* It looks like the linear fit (red line is way off, but the blue and green lines are better.
* I'll isolate them and print the log.
```{r LogPlots}
plot.co2(y=log(co2$trend),pch=19,type="l")
lines(xx, log(predict(fit1, data.frame(x=xx))), col="red")
lines(xx, log(predict(fit2, data.frame(x=xx))), col="green")
lines(xx, log(predict(fit3, data.frame(x=xx))), col="blue")
```

* I'll look at the residuals after subracting the fits - that should be the non-periodic forcings. 
```{r Residuals}
# now, look at the residuals after subtracting fits.
xFramed <- data.frame(x=co2$decimal_date)
rp2=y-predict(fit2, xFramed)
rp3=y-predict(fit3, xFramed)

#points(x=x,y=yp3,col="green")
plot.co2(y=log(rp2+1.6), col=alpha("red",.5), type="l", lwd=5)
lines.co2(y=log(rp3+1.6), col=alpha("green",.2), lwd=3)

```

* There doesn't seem to be any difference between the square and cube fits
 we will just use the square fit.
* I would like to subract the residuals, and perform a new fit, to see if the data cleans up anymore
* remove the residuals from the original data, then re-run the simulation
```{r, SubResiduals}
trr<- co2$trend - rp2
fit2b <- lm(trr~poly(co2$decimal_date,2,raw=TRUE))
rp2b = co2$trend-predict(fit2b, xFramed)
plot.co2(y=log(rp2b+1.6), type="l")
#lines(x=co2$decimal_date, y=log(rp2+1.6), col=alpha("red",.5))

plot.co2( y=co2$trend, type="l", lwd=8, col="blue")
lines.co2(y=co2$trend-rp2, type="l", lw=5, col="green")
lines.co2(y=co2$trend-rp2b, type="l", lw=3, col="red")

```

* See if there is any significance to square or cubic fit
```{r FitTest}
anova(fit1,fit2)
anova(fit1,fit3)

```

* That looks like some real activity. There is definitely a response to the eruptions, but it's subtle.
* try a spline fit to the residuals
```{r DiffEqs}
dds <- sm.spline(x, rp2b)
summary(dds)
ddyb <- predict(dds, x, 1)
summary(ddyb)
#plot(x, yp2, type='l', col="red")
plot.co2(y=ddyb, type='l', col="blue")

plot.co2(y=log(ddyb+4), type='l', col="blue")
```

